Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-14 Ben_Manuscripts/hdphmm/Draft.tex
Saving significant progress on the mechanistic front

  \textit{Relating Parameters to Solute-Membrane Interactions}: Most of what we 
  have learned thus far has been based on the parameters of our model with some 
  qualitative validation using a single methanol trajectory. While 
  we may be able to further corroborate our arguments by analyzing more
  individual trajectories, it is more useful to correlate the state behavior with
  solute-membrane interactions.

  %BJC2: A lot of the following will probably be reworked with different clustering scheme
%  \textit{Hydrogen Bonding Patterns}: 
  The types of hydrogen bonding undergone by solutes is dependent on radial
  location (see Figure~\ref{fig:hbond_pichart}). States with radial means near 1 nm
  or less (states 1, 3 and 5) typically hydrogen bond with the monomer carboxylate
  groups and ether linkages. States with radial means greater than 1 nm (states 2,
  4, 6 and 7) tend to hydrogen bond with the monomer tails. Our previous work has
  shown that there is a gradual transition from the hydrophilic to hydrophobic region in
  these membranes, meaning the monomer location is not tightly bound in $r$. 
  %MRS: above logic is not clear?
  Therefore, there are instances of hydrogen bond interactions that one might consider 
  uncharacteristic based on the state's radial mean.
  %MRS: what is implication of ``uncharacteristicness''?

  States which exhibit longer dwell times have longer hydrogen bond lifetimes (see
  Figure~\ref{fig:hbond_pichart}). States 3--7 have significantly longer hydrogen 
  bond lifetimes than states 1 and 2. It makes physical and mathematical sense that
  states 1 and 2 should not stay trapped by hydrogen bonding for long periods because
  they typically take large hops as communicated by their large covariances.
  In the tail states, hydrogen bond lifetimes appear to increase with their dwell 
  times. However, in the pores, the same linear relationship does not hold.
  For example, state 5 has the lowest hydrogen bond lifetime of states 3--7
  despite having the second longest dwell times.

  The local density experienced by solutes may have an impact on the length of solute 
  entrapment. On average, solutes in state 5 are surrounded by a relatively high density
  of particles. Based on its radial location and preference towards hydrogen bonding
  with the ether linkages (see Figure~\ref{fig:hbond_pi_charts}), it is reasonable to
  assume that while in state 5, methanol spends most of its time just outside the 
  pore region, sandwiched between the monomer head groups. Although state 5 hydrogen
  bonds break more frequently than one might expect for a state with long dwell times,
  it is possible that the solute's dense surroundings prevent it from wandering off 
  before the hydrogen bonding interaction can be recovered. This may explain the 
  behavior of the methanol in state 5 of the trajectory in Figure~\ref{fig:mechanism_map}
  where the solute's center of mass appears to move more freely about its mean than 
  other trapped states. The data suggests that methanol molecules in state 3 lie closer
  to the pore center than state 5 because they tend to hydrogen bond most frequently 
  with the carboxylate groups. Methanol molecules which experience state 3 are subject 
  to a lower density local environment and rely more heavily on hydrogen bonding interactions, 
  evidenced by their longer lifetimes, to stay trapped.

  States describing behavior in the tail region show a direct relationship between 
  hydrogen bond lifetimes and local density. Methanol molecules which experience state 2 
  spend their time in a low density region of the tails where they can easily break free
  of hydrogen bonds. Conversely, methanol molecules trapped in state 7 are in the densest 
  region of all states studied. It is likely difficult for methanol to escape these high 
  density regions and its position is further stabilized via hydrogen bonding.

  %BJC4: I didn't put sodium ion association in this figure because it's infrequent. 
  %BJC4: If I compare to acetic acid or urea, I could add it back. I was thinking of splitting
  % the pi chart in half with slices corresponding to hbonds on one side and association on the other
  \begin{figure}
  \centering
  \begin{subfigure}{0.54\textwidth}
  \includegraphics[width=\textwidth]{hbond_pi_charts.pdf}
  \caption{}\label{fig:hbond_pi_charts}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{hbond_lifetimes_density_MET.pdf}
  \caption{}\label{fig:hbond_lifetimes_density_MET}
  \end{subfigure}
  %MRS: legend doesn't show stripes.
  \caption{(a) Methanol hydrogen bonds in different segments of the LLC monomer
  dependent on the state. States with mean radial locations in the pore region tend to
  hydrogen bond with the head groups and ether linkages. States with mean radial
  locations in the tails tend to hydrogen bond with the tail oxygen atoms.
  (b) Hydrogen bond lifetimes and the local density experienced by methanol
  molecules can help us to understand dynamical states in terms of membrane 
  interactions. The bars are colored according to match the states in Figures~\ref{fig:mechanism_map}
  and~\ref{fig:common_states_MET}. States 1 and 2, which have large covariances, also have short
  hydrogen bond lifetimes and exist in relatively low density regions. The
  hydrogen bond lifetimes of methanol in states with radial means in the tail
  region tend to increase with local density and mean dwell times. 
  %MRS7: not clear from graphic which ones are in the pore.
  %MRS7: is there any legend for the color for b? If same as above figure, indicate. Don't REALLY need color since they bars are already labeled, so it's 2 sources of duplicate information.
  %BJC7: The state numbers are all you need to compare to the previous figure BUT the color make it easier 
  %plot of the methanol trajectory where states aren't labeled.
  %MRS7: cross-hatching doesn't appear in the legend? Not sure which is which (I assume hatched is number density because closer to that axis).
  %BJC7: Is it because the hatches aren't dense enough? I see them in the legend
  This monotonic
  relationship does not exist for states in the pore region. The high 
  local density experienced by state 5 methanol molecules may explain why 
  hydrogen bond lifetimes are short but dwell times are high. State 5 methanol's
  preference towards hydrogen bonds with ether linkages suggests that the solute
  sits between LLC monomer head group which prevent it from wandering to far
  before recovering its previous hydrogen bond.
  }\label{fig:hbond_pichart}
  \end{figure}

>>>>

  \textit{Relating Parameters to Solute-Membrane Interactions}: Based on the
  molecules and ions in the system, it is straightforward to hypothesize two types
  interactions that might occur.
  \begin{itemize}  
    \item All of the solutes are polar and have hydrogen bond donating and 
    accepting groups. 
    \item The monomers have 10 oxygen atoms distributed from head to tail that are 
    capable of accepting hydrogen bonds. 
    \item Also, the ionic bond between the head groups and sodium ions can
    be easily broken since both components can be stabilized by surrounding 
    water molecules and solutes.
    \item Therefore, it is logical to search for hydrogen bonding interactions
    and solute assocation with sodium ions.
  \end{itemize}

  For methanol, hydrogen bonding is a much more dominant interaction than sodium
  ion association. 
  \begin{itemize}
    \item In Figure~\ref{fig:hbond_fractions}, we show that methanol participates
    in hydrogen bonds to various parts of the monomer between 27 and 45 \% of the
    total time it occupies the dominant states.
    \item Methanol associates with sodium ions less than 10 \% of the time in 
    dominant states.
  \end{itemize}

  The distribution of types of hydrogen bonds and associated hydrogen bond lifetimes
  varies by dominant state.
  \begin{itemize}
    \item States 1 and 2 are dominated by hydrogen bonds to the head groups and, to 
    a smaller extent, the ether linkages.
    \item State 1 methanol molecules that donate hydrogen bonds to the tail oxygen 
    atoms exhibit by far the longest hydrogen bond lifetimes among all states.
    \item States 3 and 4 hydrogen bond to all regions of the monomer relatively 
    evenly, which is consistent with our earlier observation that states 3 and 4
    are representative of behavior seen radially throughout the membrane pores.
    \item State 3 has the lowest hydrogen bond and sodium association lifetimes 
    which is consistent with its short dwell times. These are fleeting interactions
    since it is likely a jumping state.
    \item Methanol molecules in state 5 hydrogen bond almost exclusively to the 
    carboxylate head groups and ether linkages which is consistent with the radial
    ean of the cluster.
  \end{itemize}

  \begin{figure}
  \centering
  \begin{subfigure}{0.475\textwidth}
  \centering
  \includegraphics[width=0.6\textwidth, angle=0]{monomer_oxygens.png}
  \caption{}\label{fig:monomer_oxygens}
  \end{subfigure}
  \begin{subfigure}{0.475\textwidth}
  \includegraphics[width=\textwidth]{hbond_fractions.pdf}
  \caption{}\label{fig:hbond_fractions}
  \end{subfigure}
  \begin{subfigure}{0.475\textwidth}
  \includegraphics[width=\textwidth]{hbond_lifetimes.pdf}
  \caption{}\label{fig:hbond_lifetimes}
  \end{subfigure}
  % BJC4: I put this one here for compact-ness. Not sure if I should separate into own figure
  \begin{subfigure}{0.475\textwidth}
  \includegraphics[width=\textwidth]{association_fraction_lifetimes.pdf}
  \caption{}\label{fig:association_fraction_lifetimes}
  \end{subfigure}
  \caption{(a) We analyzed hydrogen bond donation to the highlighted oxygen atoms in three 
  regions of the monomer: the carboxylate groups (blue), the ether linkages (green) and the
  tails (red). (b) For each dominant state, we plotted the fraction of the total time in
  that state spent hydrogen bonding to various regions of the monomer, color coded to match (a).
  (c) For each type of hydrogen bond in each dominant state, we calculated the 95th percentile
  of hydrogen bond lifetimes. (d) We also analyzed the frequency and lifetimes of solute 
  association with sodium ions. Methanol associates with sodium less than 10 \% of the total 
  time it occupies any of the dominant state.
  %Relative to hydrogen bonding, it is a relatively infrequenct interaction
  %for methanol molecules, but the 
  }
  \label{fig:hydrogen_bonding}
  \end{figure}

  The local number density of heavy atoms surrounding solutes help round out methanol's
  mechanistic picture.
  \begin{itemize}
    \item Although hydrogen bonding interactions are relatively frequent, methanol molecules
    are unbound by any measurable electrostatic interactions about half or more of the 
    simulation time.  
    \item Figure~\ref{fig:local_densities} helps to illustrate how the local density of 
    membrane components can aid in solute entrapment.
    \item When solutes experience long trapping times, the local number density of heavy 
    atoms is larger.
    \item Even when hydrogen bonds break, surrounding alkane density likely helps keep
    the solutes in place allowing them to eventually reform. This can happen deep in the
    tails or close to the monomer head groups.
    \item Methanol molecules which experience State 3 have by far the smallest local 
    number density which is likely responsible for their mobility, as suggested by the state's
    large covariance.
  \end{itemize}

  \begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{local_densities.pdf}
  \caption{We measured the local number density of heavy atoms surrounding solutes in each
  state. In blue, we plot the local number density experienced by solutes averaged over all
  frames in which the solute occupied the given state. We broke down this overall average 
  number density into number densities measured from instances when the state is occupied 
  for long (orange) versus short(red) dwell time states. We define define `long' dwells times
  as the 50th percentile of all dwell times in the given state, and short as all others. It
  is clear that solutes which experience longer dwells times are also surrounded by more 
  heavy atoms on average.}\label{fig:local_densities}
  \end{figure}

%BJC4: the following has pretty all been replaced.
%  Most of what we 
%  have learned thus far has been based on the parameters of our model with some 
%  qualitative validation using a single methanol trajectory. While 
%  we may be able to further corroborate our arguments by analyzing more
%  individual trajectories, it is more useful to correlate the state behavior with
%  solute-membrane interactions.
%
%  %BJC2: A lot of the following will probably be reworked with different clustering scheme
%%  \textit{Hydrogen Bonding Patterns}: 
%  The types of hydrogen bonding undergone by solutes is dependent on radial
%  location (see Figure~\ref{fig:hbond_pichart}). States with radial means near 1 nm
%  or less (states 1, 3 and 5) typically hydrogen bond with the monomer carboxylate
%  groups and ether linkages. States with radial means greater than 1 nm (states 2,
%  4, 6 and 7) tend to hydrogen bond with the monomer tails. Our previous work has
%  shown that there is a gradual transition from the hydrophilic to hydrophobic region in
%  these membranes, meaning the monomer location is not tightly bound in $r$. 
%  %MRS: above logic is not clear?
%  Therefore, there are instances of hydrogen bond interactions that one might consider 
%  uncharacteristic based on the state's radial mean.
%  %MRS: what is implication of ``uncharacteristicness''? 
%
%  States which exhibit longer dwell times have longer hydrogen bond lifetimes (see
%  Figure~\ref{fig:hbond_pichart}). States 3--7 have significantly longer hydrogen 
%  bond lifetimes than states 1 and 2. It makes physical and mathematical sense that
%  states 1 and 2 should not stay trapped by hydrogen bonding for long periods because
%  they typically take large hops as communicated by their large covariances.
%  In the tail states, hydrogen bond lifetimes appear to increase with their dwell 
%  times. However, in the pores, the same linear relationship does not hold.
%  For example, state 5 has the lowest hydrogen bond lifetime of states 3--7
%  despite having the second longest dwell times.

%  The local density experienced by solutes may have an impact on the length of solute 
%  entrapment. On average, solutes in state 5 are surrounded by a relatively high density
%  of particles. Based on its radial location and preference towards hydrogen bonding
%  with the ether linkages (see Figure~\ref{fig:hbond_pi_charts}), it is reasonable to
%  assume that while in state 5, methanol spends most of its time just outside the 
%  pore region, sandwiched between the monomer head groups. Although state 5 hydrogen
%  bonds break more frequently than one might expect for a state with long dwell times,
%  it is possible that the solute's dense surroundings prevent it from wandering off 
%  before the hydrogen bonding interaction can be recovered. This may explain the 
%  behavior of the methanol in state 5 of the trajectory in Figure~\ref{fig:mechanism_map}
%  where the solute's center of mass appears to move more freely about its mean than 
%  other trapped states. The data suggests that methanol molecules in state 3 lie closer
%  to the pore center than state 5 because they tend to hydrogen bond most frequently 
%  with the carboxylate groups. Methanol molecules which experience state 3 are subject 
%  to a lower density local environment and rely more heavily on hydrogen bonding interactions, 
%  evidenced by their longer lifetimes, to stay trapped.
%  States describing behavior in the tail region show a direct relationship between 
%  hydrogen bond lifetimes and local density. Methanol molecules which experience state 2 
%  spend their time in a low density region of the tails where they can easily break free
%  of hydrogen bonds. Conversely, methanol molecules trapped in state 7 are in the densest 
%  region of all states studied. It is likely difficult for methanol to escape these high 
%  density regions and its position is further stabilized via hydrogen bonding.
%  %BJC: I didn't put sodium ion association in this figure because it's infrequent. 
%  %BJC: If I compare to acetic acid or urea, I could add it back. I was thinking of splitting
%  % the pi chart in half with slices corresponding to hbonds on one side and association on the other
%  \begin{subfigure}{0.54\textwidth}
%  \includegraphics[width=\textwidth]{hbond_pi_charts.pdf}
%  \caption{}\label{fig:hbond_pi_charts}
%  \begin{subfigure}{0.45\textwidth}
%  \includegraphics[width=\textwidth]{hbond_lifetimes_density_MET.pdf}
%  \caption{}\label{fig:hbond_lifetimes_density_MET}
%  %MRS: legend doesn't show stripes.
%  \caption{(a) Methanol hydrogen bonds in different segments of the LLC monomer
%  dependent on the state. States with mean radial locations in the pore region tend to
%  hydrogen bond with the head groups and ether linkages. States with mean radial
%  locations in the tails tend to hydrogen bond with the tail oxygen atoms.
%  (b) Hydrogen bond lifetimes and the local density experienced by methanol
%  molecules can help us to understand dynamical states in terms of membrane 
%  interactions. The bars are colored according to match the states in Figures~\ref{fig:mechanism_map}
%  and~\ref{fig:common_states_MET}. States 1 and 2, which have large covariances, also have short
%  hydrogen bond lifetimes and exist in relatively low density regions. The
%  hydrogen bond lifetimes of methanol in states with radial means in the tail
%  region tend to increase with local density and mean dwell times. 
%  %MRS7: not clear from graphic which ones are in the pore.
%  %MRS7: is there any legend for the color for b? If same as above figure, indicate. Don't REALLY need color since they bars are already labeled, so it's 2 sources of duplicate information.
%  %BJC7: The state numbers are all you need to compare to the previous figure BUT the color make it easier 
%  %plot of the methanol trajectory where states aren't labeled.
%  %MRS7: cross-hatching doesn't appear in the legend? Not sure which is which (I assume hatched is number density because closer to that axis).
%  %BJC7: Is it because the hatches aren't dense enough? I see them in the legend
%  This monotonic
%  relationship does not exist for states in the pore region. The high 
%  local density experienced by state 5 methanol molecules may explain why 
%  hydrogen bond lifetimes are short but dwell times are high. State 5 methanol's
%  preference towards hydrogen bonds with ether linkages suggests that the solute
%  sits between LLC monomer head group which prevent it from wandering to far
%  before recovering its previous hydrogen bond.
%  }\label{fig:hbond_pichart}
